---
title: "Intro to R"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Female Code Lab - Greek Girls Code"
date: "15/12/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme:
      bg: "#202123"
      fg: "#B8BCC2"
      primary: "#EA80FC"
      secondary: "#00DAC6"
      base_font:
        google: Prompt
      heading_font:
        google: Proza Libre
       
---

```{r setup, include=FALSE}
if (requireNamespace("thematic")) 
  thematic::thematic_rmd(font = "auto")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy('')
```

# Welcome   

Welcome to the Female Code Lab "Intro to R" hands on tutorial delivered by Greek Girls Code.

RStudio is an Integrated Development Environment for R. What does that mean? Well, if you think of R as a language, which it is, you can think of RStudio as a program that helps you write and work in the language. RStudio makes programming in R much easier and I suggest that you use it!

**The tutorial is designed to help you: **

- Familiarize with R and RStudio IDE
- Familiarize with the types of data
- Learn R base functions
- Intro to data transformation with `tidyverse` R package


# Before we start - Participant preparation

To maintain a smooth and timely flow during the "Intro to R" section of the Female Code Lab, each participant must install and set up R, Rstudio and the tidyverse R package before the tutorial. 

To do so please follow the instructions bellow:

## How to install R


To install R on your computer you can also watch the following video at 
<https://vimeo.com/203516510>

## How to install RStudio

<https://vimeo.com/203516968>

## How to install `tidyverse` R package. 

In this section you will learn how to install `tidyverse` but the process is essentially the same for any R package <https://vimeo.com/203516241>. 

`tidyverse` [`{tidyverse}` package](https://www.tidyverse.org/). is compilation or collection of R packages each one designed to transform and visualize data. All `tidyverse` packages implement the same syntax/ grammar in the command line. 

To install call the following R function:

```{r, eval=FALSE, class.source='klippy'}
install.packages("tidyverse")
```

# Intro to R
*this section includes content from the joint ????????-Greek Girls Code workshop created and delivered by Eirini Tsirvouli* 
<https://eirinits.github.io/>


## Basic R commands and functionalities

You can use R to perform basic math just as you would do using a calculator!
You can just add numbers and get the result
```{r Addition, echo=TRUE}
#add numbers
1+2

#raise to power 6
4^6

#create a sequence of numbers (integers) from 1 to 100
seq(1,100)
```

You can use all arithmetic operators for practice 

- Try `+`, `-`, `*`, `/ ` 
- Try more complex math e.g. (2+4)^2/5 

### Logical Tests

You can use R to test logical arguments
for example 
```{r echo=TRUE}
# test if 
1/2 == 0.5

# or test if 
1/2 > 0.5
```

### Data types 

<https://vimeo.com/220490241>


### Data Structures

A vector is the most basic type of data structure in R. It is essentially a collection of data assigned to a symbol using the assignment operator <- .
```{r echo=TRUE}
# you can assign a single number to a vector x
x<-1
#then call the vector and get the result
x
# you can assign 1 dimensional data such as string of values
weight<-c(50, 60, 65, 72, 82, 94)
#call weight
weight
# you can also get the vector length (number of values contained in the vector)
length(weight)
# by calling the mode() command you can find out what kind of data the vector holds
mode(weight)
```


You can create 2 dimensional vectors by assigning other type of data structure like a **data frame**. A **data frame** is a 2-dimensional table-like structure which contains columns and rows with variables and values.

```{r echo=TRUE}
#call the mtcars data
mtcars
#find out its structure type (class)
class(mtcars)
# assign the data frame to a vector
cars<-mtcars
#call cars
cars
```

Other data structures include:

- lists , 1-D structure
- matrices, 2-D structures of homogeneous data
- arrays, multiple dimensions of homogeneous data

### Contructing and manipulating vectors and data frames: R base functions

We can construct a data.frame or a 1-D vector and extract information based on desired criteria.

```{r echo=TRUE}
# construct a 1-D vector
treatment_dose <- c("90mg","40mg","20mg","90mg","20mg")
# get the type 
mode(treatment_dose)
# get the 1st value in the sequence of values
treatment_dose[1]

# we can add a new value to the vextor

treatment_dose_plus<- c(treatment_dose, "25mg")

treatment_dose_plus

# or we can subtract a value from the vector

treatment_dose_minus<-treatment_dose[-1]

treatment_dose_minus
```

or we can assign a data frame to a vector and extract information based on our predefined criteria

```{r echo=TRUE}
#get the dimensions of mtcars
dim(mtcars)
# get the first 6 rows of mtcras
head(mtcars)
# get the column names
colnames(mtcars)
#get the rownames of mtcars
rownames(mtcars)

#subset the first 2 columns of the dataframe
mt.sub1<-mtcars[,1:2]
mt.sub1
#subset the 1st 2 rows and the first 2 columns
mt.sub2<-mtcars[1:2, 1:2]
mt.sub2
```

### Transforming and filtering a dataframe with `tidyverse`

Tidyverse is a collection of packages essential to data science and analysis. By loading the package you'll see each package contained in the collection. We will demonstrate a couple of `dplyr` functions for mutating and filtering data from the mtcars data frame. This dataframe contains columns with numeric data. `dplyr` can mutate and filter data from heterogeneous data frames with columns of numeric and character data but this will be demonstrated in the next more advanced tutorial by Irene-Angelica Chounta. 

```{r echo=TRUE}
library(tidyverse)

#multiply mpg with cyl and add the result in a new column named prod

#first we need to transform the rownames into a column named "name"
mtcars.new<-rownames_to_column(mtcars, var="name")
head(mtcars.new)

mtcars.mut<-mtcars.new %>% mutate (name, prod = (mpg*cyl) )
mtcars.mut
# now we want to select the name, mpg , cyl and the prod columns and remove all other columns 
mtcars.mut<-mtcars.mut %>% select(name:cyl,prod)
mtcars.mut
```


## Create an RStudio project

RStudio gives you an option to start a project where you can:

- create a directory to store your data, analysis files, variables and the project related packages
- start the analysis from where you left it last time you worked on the project
- collaborate and publish reproducible analyses in public repositories such as github




## Themed Plots {.tabset .tabset-pills}



### ggplot2

```{r}
library(ggplot2)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() + geom_smooth()
```

### lattice

```{r}
lattice::show.settings()
```

### base

```{r}
plot(pressure, col = thematic::thematic_get_option("accent"))
```
